---
title: "Introduction to Basic SNP Analysis with snpR"
author: "William Hemstrom"
date: "June 27, 2021"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Introduction to Basic SNP Analysis with snpR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
#bibliography: snpR.bib

---

```{r include=FALSE}
library(snpR); library(ggplot2); # library(kableExtra)
```

# snpR Introduction

snpR is a SNP genomic analysis package that is designed to do a range of basic genomic analyses, such as calculate observed and expected heterozygosity, linkage disequilibrium, pairwise-Fst, and so on. This package also has dimensionality reduction tools (eg. PCA, tsne, umap), data formatting conversions for other packages and programs, population structure, genetic prediction, private allele detection and more (See documentation for details). All functions are written to take similar arguments and interpret them in consistent ways. This is done through the use of the snpRdata object class, which all snpR functions take as their first argument. Nearly all functions are written to be overwrite safe: the object provided to the function should be overwritten by the output. Newly calculated statistics will simply be merged-in to the existing object.

A fundamental feature of snpR is the use of *facets*. Facets define categorical metadata features of the data, and can reference either features describing individual SNPs (such as chromosome) or individual samples (such as population). Facets are used to automatically break apart input data to calculate statistics in intuitive ways, such as by calculating observed heterozygosity in each population. Since metadata is contained in the snpRdata object, the user will never need to manually subset or loop calculations across categories/levels of a facet. Further, facets can be combined so that many calculations occur across these groups with just one command. Facets are combined using "." argument (Examples are provided below). Further information on facets is provided later in this guide. 

# Quick-start Guide

Here, we will walk through a basic analysis of the example three-spined stickleback data included in this package (stickSNPs). We will filter SNP data, make a PCA exploratory plot, use a clustering algorithm (tSNE), calculate minor allele frequencies, $\pi$, observed heterozygosity, see if SNPs are in HWE, check for private alleles, calculate pairwise FST, gaussian smooth and plot these parameters, calculate pairwise LD in two populations on one chromosome, and run through a Genome Wide Association example.

## Data import

snpR can take a variety of SNP data formats. The basic input format is a tab-delimited SNP dataset, where each row contains metadata and genotypes for a single SNP. Genotypes are noted with two characters, such as "GG" or "GA" for a G/G homozygote and a G/A heterozygote, respectively. The missing data format can be user-defined, but is "NN" by default. For example, see the stick_NN_input.txt distributed with this package.
```{r}
system.file("extdata", "stick_NN_input.txt", package = "snpR")
```

Note that this is identical to the format output by, for example, the ANGSD software package. Ensure that genotypes are not stored as factors during import. The resulting data looks like:

```{r echo=FALSE}
print(head(stickFORMATs$NN)[,1:6], row.names = F)
```

Note that snpR can also import data where genotypes are stored as four numbers (ex. "GG", "GA", and "NN" as "0101", "0102", and "0000", respectively), and as a single numeric character that represents the count of the minor allele (0 or 2 for homozygotes, 1 for heterozygotes). 

Data in the "NN" format, as shown above, can be imported directly into a snpRdata set alongside any corresponding metadata using import.snpR.data function:

```{r}
# example data is present in stickRAW 
# the first three characters of the column names are the population, so let's grab that.
# since the first three columns contain snp.metadata, we need to remove those before import!
sample_meta <- data.frame(pop = substr(colnames(stickRAW)[-c(1:3)], 1, 3), stringsAsFactors = F)

# grab our sample metadata
snp_meta <- stickRAW[,1:3]

# grab the genetic data
gen_data <- stickRAW[,4:ncol(stickRAW)]

# import, remember to remove metadata from the genotypes!
my.dat <- import.snpR.data(genotypes = gen_data, 
                           snp.meta = snp_meta, 
                           sample.meta = sample_meta)

```
Note: 

* Column names in the snp and sample metadata must be unique (no repeats across *both* metadatasets), and cannot contain '.', '~', or whitespace!

* Some functions expect to see columns named 'chr' and 'position'. Ensuring snp.meta has these columns early on saves some potential column renaming later.   

* Shared metadata levels across different columns can cause issues if statistics are calculated for both of the two categories (for example, if both a population and a family are both named "A").

* If you have data in the *adegenet* genind or genlight formats, these can also be automatically imported using import.snpR.data! There are examples for this in the import.snpR.data function documentation. More common formats are planned for future support. 

Importing data into snpR can also be done using the format_snps function. This function takes a few extra arguments. First, it takes the data and a note of the input format, then it needs information on the number of snp metadata columns, the missing data notation, and any sample metadata.

```{r}
# Import numeric formatted data.
head(stickFORMATs$`0000`)[,1:6]

# get sample metadata, as before.
sample_meta <- data.frame(pop = substr(colnames(stickFORMATs$`0000`)[-c(1:4)], 1, 3), stringsAsFactors = F)

# import
my.dat2 <- format_snps(stickFORMATs$`0000`, input_format = "0000", input_meta_columns = 4, input_mDat = "0000", sample.meta = sample_meta)
```

The result will be the same as import.snpR.data! There are a handful of other possible data imports that format_snps can handle, these are described in the function's documentation.

## snpRdata objects and facets

### snpRdata objects
snpRdata objects store a range of variables internally in order to minimize repeat computation of things like allele frequencies.

Calling a snpRdata object will show a summary of the calculations done on the object and the metadata categories available.

```{r}
#my.dat #except for when knitting this document, apparently
```

By and large, snpRdata objects should be accessed via accessor functions:
```{r}

# get the number of samples
nsamps(my.dat)
ncol(my.dat)

# get the number of SNPs
nsnps(my.dat)
nrow(my.dat)

# dim also works
dim(my.dat)

# view sample metadata
head(sample.meta(my.dat))

# write to sample metadata; note that writing to any part of the metadata or genotypes will cause any calculations
# to be discarded, since the new columns can cause some unexpected results.
sample.meta(my.dat)$fam <- sample(LETTERS[1:4], nsamps(my.dat), T)

# view snp metadata
head(snp.meta(my.dat))

# write to sample metadata; note that writing to any part of the metadata or genotypes will cause any calculations
# to be discarded, since the new columns can cause some unexpected results.
snp.meta(my.dat)$new_snp_col <- sample(LETTERS[1:6], nsnps(my.dat), T)

# view genotypes, can be written to as above
genotypes(my.dat)[1:6, 1:6]

```


Almost all snpR functions are built to allow overwriting for easy object management. Since calculated data is stored in snpRdata objects, snpR functions generally return the snpRdata object that was given to them, but with extra data merged in. For example, to calculate observed heterozygosity, the calc_ho function can be used:

```{r}
# calculate observed heterozygosity, overwriting the my.dat object
my.dat <- calc_ho(my.dat)

```

Calculated statistics can be fetched using the get.snpR.stats accessor function. Note that this function returns a data.frame or nested list (depending on the requested stats), not a snpRdata object, and *should never be used to overwrite a snpRdata object!*. Since some calculations are run pairwise you need to specify if you want the pairwise or single stats returned. 

```{r}
# fetch calculated statistics
ho <- get.snpR.stats(my.dat, stats = "single")
head(ho)

```


### Facets

Facets are a fundamental feature of snpR. Often, SNP data must be separated by population/chromosome/family/year etc. before statistics are calculated. Additionally, many statistics, like Fst, need to be calculated *pairwise* by comparing each population to every other population. Facets allow you to easily do this without the need for direct data manipulation. 

#### Basic Facets
Facets are defined by reference to column names in either the snp or sample metadata that was provided when creating a snpRdata object. For example, the facet "pop" would split my.dat apart by the pop column in the sample metadata and run the requested function or calculation for each of the levels/subfacets within that facet (eg. the facet pop comprises 6 levels/subfacets - ASP, CLF, OPL, PAL, SMR, UPD). Whereas the facet "group" would split my.dat apart by the "group" column in the snp metadata (21 levels/subfacets) and run any specified calculation for each unique level/subfacet. 

To calculate observed heterozygosity at each SNP within each population, for example, one can supply the "pop" facet.

```{r}
my.dat <- calc_ho(my.dat, facets = "pop") #running the calculation
head(get.snpR.stats(my.dat, facets = "pop", stats = "single")) #retrieving the results

```

#### Complex Facets

Facets can define more than one column at once, in which case the data is split by *both* columns! Within a single facet, column names are separated by a '.'.  For example, if there was both a pop and fam (or family) column in the sample metadata, the facet "pop.fam" would split by both the pop and the fam columns simultaneously, so it would have categories like "popA.famB" for an individual from population A, family B. These "complex" facets can also note both SNP and sample-specific categories simultaneously, for example, "pop.group" would split the data by the pop column in the sample metadata and the group column in the SNP metadata simultaneously. 

```{r}
my.dat <- calc_ho(my.dat, "pop.fam") #running the calculation
head(get.snpR.stats(my.dat, "pop.fam", stats = "single")) #retrieving the results
```

The order of the categories given in a facet do not matter--"pop.group" will produce the same result as "group.pop".

#### Multiple facets

Most functions can also take several facets at once, given as a character vector. For example, to split by both population, then by population and group simultaneously, c("pop", "pop.group") could be given.


```{r}
my.dat <- calc_ho(my.dat, c("pop.fam", "pop"))
head(get.snpR.stats(my.dat, c("pop.fam", "pop"), stats = "single")) # will show both facets, but it's a pretty large dataset.
tail(get.snpR.stats(my.dat, c("pop.fam", "pop"), stats = "single")) # both facets are contained in the dataset.
```

A handful of functions only work with either SNP or sample specific facets, but this is fixed "under the hood." For example, since observed heterozygosity is calculated independently for each SNP, it makes no sense to calculate it split up by chromosome. Using the "pop.group" facet with calc_ho, therefore, will internally default to "pop".


```{r}
# same as just doing "pop" as a facet.
my.dat <- calc_ho(my.dat, "pop.group") 
head(get.snpR.stats(my.dat, "pop.group", stats = "single"))
```

Most of the time, giving NULL (the usual default) to a facets argument will run the data without splitting it up at all. Resulting stats will belong to the ".base" facet, a special, internally used facet that implies no splitting! Giving "all" to a facets argument, on the other hand, will run the data splitting by all previously run facets!

```{r}
# same as just doing "pop" as a facet.
my.dat <- calc_ho(my.dat) 
head(get.snpR.stats(my.dat, ".base", stats = "single"))
```

## SNP filtering

The filter_snps function can be used to filter poorly sequenced individuals and loci, remove SNPs with a low minor allele frequency either overall or by facet (often population), remove highly heterozygous SNPs (which are likely sourced from duplicated genomic regions), and remove non-biallelic SNPs.

Here, we will filter out all loci with a minor allele frequency below 0.05 in all populations, loci with a heterozygote frequency above 0.55, that are not bi-allelic, that are non-polymorphic, or that are sequenced in less than half of the individuals. We will also remove all individuals that are genotyped at less than half of the remaining loci, then quickly re-check that all loci are still polymorphic.


```{r, results = "hide"}
my.dat <- filter_snps(x = my.dat,
                         maf = 0.05, # what is the minimum minor allele frequency to keep?
                         hf_hets = 0.55, # what is the maximum heterozygote frequency to keep?
                         min_ind =  0.5, # remove all loci sequenced at less than half of the individuals.
                         min_loci = 0.5, # remove all individuals sequenced at less than half of the loci
                         re_run = "partial", # tell filter_snps to recheck that all loci are polymorphic after removing individuals. Could set to "full" to re-check maf and heterozygote frequency, or FALSE to skip.
                         maf.facets = "pop", # we want to keep any SNPs that pass the maf filter in ANY population!
                         non_poly = TRUE, # we want to remove non-polymorphic loci
                         bi_al = TRUE) # we want to remove non bi-allelic loci") # missing genotypes are listed as "NN"
```

Note that we use the facet notation defined above to define maf.facets.

## PCA plots

A common first analytical step for new data is to run a PCA to look for any evidence of population structure/sequencing errors. snpR has a function to do this in a streamlined way. We'll also make a t-stochastic neighbor embedding (tSNE) plot as well. tSNE is a machine learning method that compresses more PCA axes onto a single two dimensional plot, and can be really nice for showing higher-order structure such as family groups within the sequenced individuals. Another algorithm uniform manifold approximation and projection (UMAP) can be used with the plot_clusters function call, but we will not be using umap for these examples. 

Note 

* A PCA cannot be constructed with missing data, so we interpolate with the average number of each allele across all individuals as the default ("bernoulli"). There are other interpolation methods available in this package (see plot_clusters documentation for more details). 
* Filtering data to remove poorly sequenced individuals and loci before PCA is recommended. 
* tSNE and UMAP require installation of additional libraries in order to run (see documentation for more details).

Both the PCA and tSNE plotting functions can work with a column of population IDs (as well as one additional info column, such as sex). The stickSNPs dataset has an additional sample metadata column called "fam", which relates to family, so we'll use that. If you recall, "fam" was created earlier for my.dat in the snpRdata objects section.  

```{r}
head(sample.meta(my.dat))
```

To run a PCA and tSNE, we can use the plot_clusters function. Note that this function, unlike all other snpR plotting functions, **does not return a snpRdata object** and we should **not overwrite our data!** so, save it to a new variable. If you do not supply a plot_type argument all possible plot types will run (PCA, tSNE, UMAP) which can be time and computationally intensive depending on the dataset. 


```{r, results='hide'}
# generate both tSNE and PCA plots
p <- plot_clusters(my.dat, facets = c("pop.fam"), plot_type = c("pca", "tsne"))
```

To view the plots, we can call two objects under the "plots" element in the list that plot_clusters returned.

```{r}
# View the PCA
p$plots$pca

```

```{r}
# view the tSNE
p$plots$tsne
```

plot_clusters also returns the raw plot data under the data element.

There are other parameters that can also be tweaked in the t-SNE function, such as perplexity and theta. For more information, check out the documentation for Rtsne::Rtsne.

Tip: you can specify viridis palette option for your plot! Or, if you want more customization, save the plot to a new object and use your preferred plotting library package. For example:

```{r, echo=TRUE, eval=FALSE}
a <- p$data$pca
ggplot2::ggplot(a, aes(PC1, PC2, color=pop))+geom_point() # a simple example

```
## Calculate minor allele frequencies, $\pi$, Ho, check for private alleles and HWE.

Minor allele frequencies, $\pi$, Ho, the number of private alleles, and HWE divergence, and other basic diversity statistics are easy to calculate with snpR. Each use a very similar function, which takes identical arguments. Since each returns a copy of the input snpRdata object, but with new information merged in, we can safely overwrite the original object each time we use these functions.

Each of these functions, like most functions, take a snpRdata object and facets as their first two arguments. Here, we'll use the my.dat object and calculate various statistics splitting the dataset by fam and pop.

```{r}
# calculate minor allele frequencies, pi, ho, check for private alleles and HWE.
my.dat <- calc_maf(my.dat, "fam.pop")
my.dat <- calc_pi(my.dat, "fam.pop")
my.dat <- calc_ho(my.dat, "fam.pop")
my.dat <- calc_private(my.dat, "fam.pop")
my.dat <- calc_hwe(my.dat, "fam.pop")

```

We can then use get.snpR.stats to view the results!

```{r}
stats <- get.snpR.stats(my.dat, "all", stats = "single") # using the "all" facet requests statistics for all facets that have been run. Specify that the returned stats are of the "single" type and not a different kind (eg. "pairwise", "association")

head(stats)
```

## Pairwise Fst calculation

snpR can calculate pairwise Fst using a few approaches. Here, we'll use Wier and Cockerham (1984)'s method to calculate pairwise Fst between each pair of populations for each SNP. 

```{r}
my.dat <- calc_pairwise_fst(my.dat, "pop", method = "WC")
```

To recover the data, we again use get.snpR.stats, but, since this Fst is a "pairwise" statistic, where each statistic is a contrast between two sample-specific facet categories, it is stored in a different slot in the snpRdata. As such, we specify that we are looking for a different kind of data.

```{r}
stats <- get.snpR.stats(my.dat, "pop", stats ="pairwise")

head(stats)
```
Note that the comparison for each Fst value is noted under the comparison column!

A full description of the different types of data we can access with get.snpR.stats is available in the documentation for that function, accessed by help(get.snpR.stats)

There are a handful of other Fst methods available in calc_pairwise_fst. Check calc_pairwise_fst documentation for details. 

## Gaussian Smoothing

snpR has a tool to make sliding window averages for parameters using a Gaussian kernel, splitting by any number of different variables. Here, we'll smooth $\pi$, Ho, and pairwise Fst, splitting by population. In addition, we'll weight the contribution of each SNP to each window by the number of times that SNP was sequenced across individuals.

To calculate sliding window averages, we need to define the sizes of the windows and the distance we slide between each. These are defined by the "sigma" and "step" arguments, respectively. Each window will include all SNPs within 3 x sigma of the window center, after which their contribution to the window average is small enough to be negligible. Both sigma and step are defined in *kilobases*.

Note that snpR looks for a column titled "position" in the SNP metadata when smoothing. This column should contain the position of each SNP in bp on the scaffold/chromosome/etc.

By default, snpR will calculate sliding window averages for all previously calculated stats. The "stats.type" argument can be used to limit this to just single ($\pi$, Ho, etc) or pairwise stats (Fst). Note that individual stats cannot be specified, since the computational gain in running only one or two stats is minimal, so only the "pairwise", "single" or c("pairwise", "single") stats.type options are allowed.

If you have many different facet combinations and a large dataset, this may take a while. To speed it up, this, as well as quite a few other functions, can be run in parallel using the par argument.


```{r, results='hide'}

ncores <- parallel::detectCores()
ncores <- ifelse(ncores == 1, 1, floor(ncores/2))
my.dat <- calc_smoothed_averages(x = my.dat, facets = "pop.group",
                                 sigma = 200, # using a window size of 200
                                 step = 50, # using a step size of 50 between windows
                                 par = ncores) # in parallel 


```

We can view the results with get.snpR.stats as usual! Since the results are a different kind of data, we view them with a different "type" argument. For the single (non-pairwise) stats, we can use the "single.window" type: 

```{r}
stats <- get.snpR.stats(my.dat, "pop.group", "single.window")

head(stats)

```

And for the pairwise windows, we use "pairwise.window".

```{r}
stats <- get.snpR.stats(my.dat, "pop.group", "pairwise.window")

head(stats)

```

## Plot the Smoothed Parameters

Next, we want to plot these variables. We'll use the ggplot2 package, and start with $\pi$:

```{r}
stats <- get.snpR.stats(my.dat, "pop.group", "single.window")

# Plot pi, pi works the same way!
ggplot(stats, aes(x = position, y = pi, color = subfacet)) + geom_line() + theme_bw() +
  facet_wrap(~snp.subfacet, scales = "free_x") # facet wrap on group to split by linkage group/chromosome.
```

Then we'll do Fst, but restrict to one linkage group to save some space.

```{r}

stats <- get.snpR.stats(my.dat, "pop.group", "pairwise.window")


ggplot(stats[stats$snp.subfacet == "groupXIX",], aes(x = position, y = fst, color = subfacet)) + geom_line() + theme_bw()
```

Note that these aren't particularly interesting plots in this case, since the example dataset has a very low SNP resolution.

## Calculate Pairwise LD

Now, we'll calculate pairwise LD for one population. We'll use Burrow's Composite Linkage Disequilibrium as our measure of LD, since it runs quickly. Several other options are also available (for more details see the documentation). We can specify that we want to split by linkage groups/chromosomes, as usual. This will produce two outputs: a proximity table and a matrix of pairwise LD for each level for each statistic.

Since LD comparisons take a while, it can be worthwhile to run only a few different categories of a given facet. The snpR LD function can do this with the "subfacet" argument. This argument takes a named list, where each entry contains the categories to run for a specific facet. For example:

```{r, results="hide"}
list(pop = "ASP", group = "groupIX") #

```
would run the ASP population for the groupIX chromosome, whereas

```{r}
list(pop = c("SMR", "OPL"), group = c("groupIX", "groupIV"))

```

would run the SMR and OPL populations for the group IX and IV chromosomes. We can use complex (multi-category) facets as normal, so

```{r}
list(fam.pop = "A.ASP")
```
would run only samples in family A and the ASP population.

In this case, let's run the ASP and OPL populations for groupIX:

```{r}
my.dat <- calc_pairwise_ld(my.dat, facets = "group.pop", 
                           subfacets =  list(pop = c("ASP", "OPL"), group = "groupIX"))

```

The results can be accessed as usual, although they are in a slightly different format!

```{r}
stats <- get.snpR.stats(my.dat, "group.pop", "LD")

str(stat)
```

The results are a multi-level named list. The first level references "prox" and "matrices". "prox" is fairly straightforward:

```{r}
head(stats$prox)
```

Each row contains information on one single SNP-SNP comparison, with information on the group and population of the comparison, as well as the positions of the SNPs, their proximity to each other, and the LD statistics.

The "matrices" object is another multi-level list. The first list level names the sample-specific facet levels, the second list level names any SNP specific facet levels, and the third list level contains the actual LD matrices. For example:

```{r}
stats$matrices$ASP$groupIX$CLD[1:10, 1:10]
```

would give us the CLD matrix for the ASP population on group IX.

Heatmaps are a convenient way to visualize LD data. snpR contains a plotting utility using ggplot for any calculated LD data.

```{r}
p <- plot_pairwise_LD_heatmap(my.dat, "group.pop")

p$plot
```

Specific plots can be requested using the snp.subfacet and sample.subfacet arguments. These will plot only the specified levels of the sample and snp-specific parts of the requested facet. For example, to plot only the ASP population:

```{r}
p <- plot_pairwise_LD_heatmap(my.dat, "group.pop", sample.subfacet = "ASP")

p$plot
```

Multiple levels can be requested by providing a vector of requested levels, for example c("ASP", "OPL"). See the documentation for plot_pairwise_LD_heatmap for more information!

## Bootstraps

Bootstraps for sliding windows can be easily calculated using the do_bootstraps function. This will create bootstrapped windows by taking real windows from the dataset and assigning random SNPs to each SNP in the window. In essence, this means that each window is a random genomic window with shuffled SNPs.

This function can do multiple statistics at once. The random windows for each statistic will be identical to save computation, but this means that each statistic is not a truly independent from the others. By default, p-values for each window based on bootstraps will be calculated and merged in to that saved statistics.

Since this can be slow, the par argument can be used to run the calculations in parallel.

```{r, results='hide'}
my.dat <- do_bootstraps(my.dat, "group.pop", boots = 1000, sigma = 200, step = 50, statistics = "all", par = 8)
```
```{r}
stats <- as.data.frame(get.snpR.stats(my.dat, "group.pop", "single.window"))
head(stats)
```

P-values can also be calculated later using the calc_p_from_bootstraps function. Note that the p-values have several different corrections. These are made using several of the multiple testing options, described in the documentation, treating the comparison family as either the entire set of calculated statistics or as just the p-values in the specific facet level.

## Easy mode

Most of the statistics above can be calculated in one step!

```{r, results='hide'}
# not run, same as above
# my.dat2 <- calc_basic_snp_stats(stickSNPs, "group.pop", sigma = 200, step = 50, par = 8)

```

This will calculate $\pi$, Ho, HWE, mafs, private alleles, and pairwise Fst, and calculate smoothed windows for these stats in one easy step!

Stats are reviewed the same way as previously documented (more details in the get.snpR.stats documentation).
```{r}
stats <- get.snpR.stats(my.dat2, "group.pop", "weighted.means")
head(stats)
```

## Genetic Association (GWAS)
Association methods include gmmat, armitage, odds ratio, and chi squared tests see the calc_association documentation for more details. Note - you may need to install GAWK in order to use the gmmat method if you are using a unix based operating system. 

```{r, results='hide'}
# stickSNPs does not have categorical phenotypic metadata so we will need to create some for this example
# add a dummy phenotype and run an association test.
x <- stickSNPs
set.seed(5465)
sample.meta(x)$phenotype <- sample(c("A", "B"), nsamps(stickSNPs), TRUE)
x <- calc_association(x, facets = c("pop"), response = "phenotype", method = "armitage")
  
```

Retrieving association statistics is similar to other stats commands (see documentation - get.snpR.stats for more details). Note if you calculate across a facet but do not request the same facet in get.snpR stats data may appear missing.
```{r}
stats <- as.data.frame(get.snpR.stats(x = x, facets = "pop", stats = "association")) 
head(stats)
```

## Pedigree Construction and visualization
snpR has a couple options for parentage assignment and pedigree construction. COLONY (Jones and Wang, 2010) is a well known program and works very well when much information is known about individuals in the sample dataset. snpR functions are designed to work with the commandline implementation of COLONY. Sequoia (Huisman, 2017) is designed to easily handle analyses with multi-generational samples.

#### COLONY
Formatting data to run in COLONY can be difficult especially with complicated datasets, but snpR can help make the process easier. The function write_colony_input takes a snpR object and writes a file that can be pulled into the commandline based COLONY interface.

write_colony_input writes a text file formatted for the COLONY commandline interface. This file is saved in a folder named 'colony' in your working directory. If the folder does not already exist, one will automatically be created. 

There are many parameters that can be specified for COLONY- check the COLONY manual and snpR documentation for more details. 

If you do not split out data into paternal, maternal, and offspring before running the write_colony_input function, the resulting file will be set for COLONY to run analyses as if all samples are potential siblings. If you have known potential parents the data can be split using the subset_snpRdata function (examples below). 

Let's go through an example to produce a simple COLONY input file. In this example all individuals would be run as potential siblings in COLONY. 

```{r, echo=TRUE, eval=FALSE}
# Note: code not run for simplicity
write_colony_input(x = stickSNPs, outfile = "sticksnps1.dat", sampleIDs = .sample.id, seed = 4562)
```

Preparing COLONY with candidate parents requires more preparation and knowledge about your dataset.
```{r, eval=FALSE, echo=TRUE}
# a more complex example requires 1) creating and adding variables to the stickSNPs example dataset and 2) creating more snpR objects.
# Note: code not run for simplicity

# first creating new variables for the dataset 
a <- 2013:2015 # possible birthyears
b <- c("M", "F", "U") # possible sexes
stk <- stickSNPs # new snpR object to manipulate
sample.meta(stk)$BirthYear <- sample(x = a, size = nsamps(stk), replace = TRUE) # create birthyears
sample.meta(stk)$ID <- 1:nsamps(stk) # create unique sampleID
sample.meta(stk)$Sex <- sample(x= b, size = nsamps(stk), replace = TRUE) # create sexes

# generating snpR objects for male and female potential parents and offspring (selecting only inds with "known" sex)
lsir <- which(stk@sample.meta$Sex =="M" & stk@sample.meta$BirthYear == "2013") # list of sires
ldam <- which(stk@sample.meta$Sex =="F" & stk@sample.meta$BirthYear == "2013") # list of dams
loff <- which(stk@sample.meta$BirthYear %in% c("2014","2015")) # list of offspring

# creating new snpR objects for Colony formatting
sir <- subset_snpR_data(x = stk, .samps = lsir) 
dam <- subset_snpR_data(x = stk, .samps = ldam)
off <- subset_snpR_data(x = stk, .samps = loff)

# ok, now ready to write colony input
write_colony_input(x = off, maternal_genotypes = dam, paternal_genotypes = sir, outfile = "sticksnps2.dat", seed = 5674) # see the documentation for write_colony_input for details on more options

```
#### Sequoia
Another option is to use Sequoia! The function run_sequoia is designed to take a snpR data object and run parentage analysis and pedigree construction. Sequoia is a separate R package, so you will need to install it in order to run these analyses. Sequoia requires metadata individual ID (ID), Sex (Sex), and either 1) BirthYear or 2) BY.min and BY.max. If you do not know the age of the sequenced individuals BY.min and BY.max are great!

Sequoia requires use of snps with high minor allele frequency (~0.3 or more), snpR objects can be filtered in the command directly! Do note that SNPs that are in fewer than 50% of individuals will be automatically dropped by sequoia. There are additional arguments and parameters that can be added in the base run_sequoia function (see the sequoia and run_sequoia documentation for more details).

Now let's go through an example of parentage and pedigree construction with sequoia.
```{r, eval=FALSE, echo=TRUE}
# Note: code not run for simplicity

# first creating new variables for the dataset 
a <- 2013:2015 # possible birthyears
b <- c("M", "F", "U") # possible sexes
stk <- stickSNPs # new snpR object to manipulate
sample.meta(stk)$BirthYear <- sample(x = a, size = nsamps(stk), replace = TRUE) # create birthyears
sample.meta(stk)$ID <- 1:nsamps(stk) # create unique sampleID
sample.meta(stk)$Sex <- sample(x= b, size = nsamps(stk), replace = TRUE) # create sexes

# data does not need to be subset to run sequoia!!
# parentage is run fairly quickly with sequoia, but sibling clusters take a longer time to run. 
# run_parents must be run prior to (or at the same time) selecting run_pedigree.
stk.ped <- run_sequoia(x = stk, facets = "pop", run_dupcheck = FALSE, run_parents = TRUE, run_pedigree = TRUE, min_maf = 0.3) # sequoia requires use of snps with high minor allele frequency

```

This command returns nested lists of outputs, for each layer of analysis and facet. Since the stickSNPs dataset does not have any parents assigned, we have made an artificial pedigree - stickPED.  

#### Visualizing your pedigrees
Assignment results from COLONY or sequoia can be plotted using a handful of packages. snpR has a function calling on visPedigree (Sheng, 2018) to make plots from COLONY BestCluster or sequoia pedigree construction outputs. 

*****need to get the function going! better
* color options
* labeling
* show all, or only inds with assigned relationships

function pedigree
* visPedigree package by (citation) and other option by kinship2 package (citation)
* visPedigree was written for use in wild animals.
* visPedigree default view is by generation, not by BirthYears



## Table of supported export formats 

snpR can export data in a wide array of commonly used formats.

```{r, echo=FALSE, results='asis'}
# tbl <- read.csv("~/R/snpR/vignettes/exportstable.csv", header = T)
# print(knitr::kable(tbl))
```
